### 2022-02-17 introduction/introduction.tex
more

A bigger breakthrough happened in 1965 when Robinson introduced a calculus that
would shape the future of automatic reasoning --- the resolution calculus
\cite{ar-65-resolution}. Consisting of a single inference rule it was simple and
elegant. It was also \emph{complete}: each theorem could be proven using this
system. Furthermore, it was less explosive than Davis and Putnam's algorithm
since it provided better guided search space exploration.

@@

Even though the successes of proving Robbins conjecture showed that it is
possible, automatically proving hard mathematical problems is still out of reach
for theorem provers. However, there are other ways in which humans and computers can work together to construct formal
proofs.

@@

We identified three stops on the road to full higher-order logic. A
distinguished feature of higher-order logic is that it not only supports
quantification over objects, but also over functions that manipulate these
objects. Thus, higher-order logic allows us to make a statement ``all functions
with arguments $0$ and $1$ result in $2$" ($\forall x. \,  x (\cst{0}, \cst{1}) \eq
\cst{2}$). Replacing $x$ with the addition function ($\cst{+}$) is enough to
disprove this statement. The first stop is to extend E to support
quantification over functions that are already present in the theory.

@@

The main contribution of this thesis is the extension of two theorem provers,
E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
extend these provers, we faced many challenges. Some of them were of engineering
nature, some of them were algorithmic, while others concerned fine-tuning the
heuristics to achieve the best possible performance. We discuss those challenges in more detail:
\begin{itemize}
  \item We implemented the complete superposition calculus for $\lambda$-free
  higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
  We had to modify internal term representation of E to support $\lambda$-free
  higher-order terms, implement new unification and matching algorithms, extend
  term indexing data structures to work with higher-order terms, and to improve the
  performance of heuristics on higher-order problems.
  \item Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second stage described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in short amount of time we decided to
  implement the calculus in Zipperposition. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It finished third, 1 percent point behind Leo-III
  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
  \cite{cb-12-satallax}.
  \item Through this experience we identified one of the bottlenecks of
  Zipperposition: its unification algorithm. To remove this bottleneck we
  developed a complete higher-order unification algorithm that was more
  efficient and generated less redundant unifiers than the state of the art. 
  \item We also noticed that Zipperposition's support for first-class Booleans
  was lacking compared to competition. We have closely studied the problems on
  which Zipperposition fails, but are easily solved by other provers and created
  a set of incomplete rules that enhance Zipperposition's Boolean support.
  Zipperposition 2 implemented these rules together with the new unification
  algorithm. In 2020, Zipperposition 2 won the higher-order division
  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points ahead of the second best
  prover, Satallax.
  \item Inspired by this success we implemented the complete superposition
  calculus for higher-order logic with Booleans (i.e., the third stage) that we developed with Bentkamp and other
  colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
  ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further fine-tuned the heuristics and implemented new ways
  to tame the search space explosion. In 2021 we released Zipperposition 2.1
  which implemented these new features. It again won
  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
  runner-up, Vampire 4.5 \cite{lkav-13-vampire}. 
  \item Taking all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic, obtaining a prover we call
  \newehoh{}. We again modified the term representation, implemented new unification and
  matching algorithms, extended indexing data structures, and tweaked
  heuristics. Even though this extension has larger scope than the original
  extension from E to Ehoh, the strong basis that we have earlier built made it
  manageable.

>>>>

A bigger breakthrough happened in 1965 when Robinson introduced a calculus that
would shape the future of automatic reasoning --- the resolution calculus
\cite{ar-65-resolution}. Consisting of a single inference rule it was simple and
elegant. It was also \emph{complete}: each theorem could be proven using this
system. Furthermore, it was less explosive than Davis and Putnam's algorithm
since it provided better guidance for search space exploration.

@@

Even though the successes of proving the Robbins conjecture showed that it is
possible, automatically proving hard mathematical problems is still out of reach
for theorem provers. However, there are other ways in which humans and computers can work together to construct formal
proofs.

@@

We have identified three stops on the road to full higher-order logic. A
distinguished feature of higher-order logic is that it not only supports
quantification over objects, but also over functions that manipulate these
objects. Thus, higher-order logic allows us to make a statement ``all functions
with arguments $0$ and $1$ result in $2$" ($\forall x. \,  x (\cst{0}, \cst{1}) \eq
\cst{2}$). Replacing $x$ with the addition function ($\cst{+}$) is enough to
disprove this statement. The first stop is to extend E to support
quantification over functions that are already present in the theory.

@@

The main contribution of this thesis is the extension of two theorem provers,
E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
extend these provers, we faced many challenges. Some of them were of engineering
nature, some of them were algorithmic, while others concerned fine-tuning the
heuristics to achieve the best possible performance. We discuss those challenges in more detail:
\begin{itemize}
  \item We implemented the complete superposition calculus for $\lambda$-free
  higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
  We had to modify the internal term representation of E to support $\lambda$-free
  higher-order terms, implement new unification and matching algorithms, extend
  term indexing data structures to work with higher-order terms, and to improve the
  performance of heuristics on higher-order problems.
  \item Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second stage described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in short amount of time we decided to
  implement the calculus in Zipperposition. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It finished third, 1 percent point behind Leo-III
  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
  \cite{cb-12-satallax}.
  \item Through this experience we identified one of the bottlenecks of
  Zipperposition: its unification algorithm. To remove this bottleneck we
  developed a complete higher-order unification algorithm that was more
  efficient and generated less redundant unifiers than the state of the art. 
  \item We also noticed that Zipperposition's support for first-class Booleans
  was lacking compared to competition. We have closely studied the problems on
  which Zipperposition fails, but are easily solved by other provers and created
  a set of incomplete rules that enhance Zipperposition's Boolean support.
  Zipperposition 2 implemented these rules together with the new unification
  algorithm. In 2020, Zipperposition 2 won the higher-order division
  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points ahead of the second best
  prover, Satallax.
  \item Inspired by this success we implemented the complete superposition
  calculus for higher-order logic with Booleans (i.e., the third stage) that we developed with Bentkamp and other
  colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
  ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further fine-tuned the heuristics and implemented new ways
  to tame the search space explosion. In 2021 we released Zipperposition 2.1
  which implemented these new features. It again won
  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
  runner-up, Vampire 4.5 \cite{lkav-13-vampire}. 
  \item Taking all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic, obtaining a prover we call
  \newehoh{}. We again modified the term representation, implemented new unification and
  matching algorithms, extended indexing data structures, and tweaked
  heuristics. Even though this extension has larger scope than the original
  extension from E to Ehoh, the strong basis that we have earlier built made it
  manageable.
